Arrows for Parallel Computations

### 2017-07-05 paper/content/arrows.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
		% \subcaptionbox
{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
		% \subcaptionbox
{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
|first| & |second| \\
\midrule
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{images/andandand}}\\
|(***)|\label{fig:***Img} & |(&&&)| \label{fig:&&&Img} \\
	\end{tabular}
	\caption{Visual depiction of syntactic sugar for Arrows.}
	\label{fig:syntacticSugarArrows}
\end{figure}
With this typeclass in place, Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): The combinators |second|, |***| and |&&&|. 
The combinator |second| is the mirrored version of |first| (Appendix~\ref{utilfns}).
The |***| function combines |first| and |second| to handle two inputs in one arrow, is defined as
% \begin{figure}[h]
\begin{code}
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{code}
% \caption{The (***) combinator}
% \label{fig:***}
% \end{figure}
while the |&&&| combinator, that constructs an arrow which outputs two different values like |***|, but takes only one input, is:
% \begin{figure}[h]
\begin{code}
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{code}
% \caption{The (\&\&\&) combinator}
% \label{fig:&&&}
% \end{figure}
A~first short example given by Hughes on how to use arrows is addition with arrows:
%|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
% \begin{figure}[h]
\begin{code}
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{code}
% \caption{Add over arrows}
% \label{fig:addArrows}
% \end{figure}
% The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
% \\\\
% \textit{Note: In the definitions Hughes gave in his paper, the notation |a b c| for an arrow from |b| to |c| is used. We use the equivalent definition |arr a b| for an arrow from |a| to |b| instead, to make it easier to find the arrow type in type signatures.}
%

The more restrictive interface of Arrows allows for more elaborate composition and transformation combinators---a monad can be \emph{anything}, an Arrow is a process of doing something, a \emph{computation}. One of the major problems in parallel computing is, however, composition of parallel processes.

>>>>

\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
		% \subcaptionbox
{\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
		% \subcaptionbox
{\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
|first| & |second| \\
\midrule
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		% \subcaptionbox
{}{\includegraphics[width = 1.5in]{images/andandand}}\\
|(***)|\label{fig:***Img} & |(&&&)| \label{fig:&&&Img} \\
	\end{tabular}
	\caption{Visual depiction of syntactic sugar for Arrows.}
	\label{fig:syntacticSugarArrows}
\end{figure}
With this typeclass in place, Hughes also defined some syntactic sugar (Fig.~\ref{fig:syntacticSugarArrows}): The combinators |second|, |***| and |&&&|. 
The combinator |second| is the mirrored version of |first| (Appendix~\ref{utilfns}).
The |***| function combines |first| and |second| to handle two inputs in one arrow, is defined as
% \begin{figure}[h]
\begin{code}
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{code}
% \caption{The (***) combinator}
% \label{fig:***}
% \end{figure}
while the |&&&| combinator, that constructs an arrow which outputs two different values like |***|, but takes only one input, is:
% \begin{figure}[h]
\begin{code}
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{code}
% \caption{The (\&\&\&) combinator}
% \label{fig:&&&}
% \end{figure}
A~first short example given by Hughes on how to use arrows is addition with arrows:
%|add| over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
% \begin{figure}[h]
\begin{code}
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{code}
% \caption{Add over arrows}
% \label{fig:addArrows}
% \end{figure}
% The benefit of using the |Arrow| typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created |add| combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple Monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
% \\\\
% \textit{Note: In the definitions Hughes gave in his paper, the notation |a b c| for an arrow from |b| to |c| is used. We use the equivalent definition |arr a b| for an arrow from |a| to |b| instead, to make it easier to find the arrow type in type signatures.}
%

The more restrictive interface of Arrows allows for more elaborate composition and transformation combinators---a Monad can be \emph{anything}, an Arrow is a process of doing something, a \emph{computation}. One of the major problems in parallel computing is, however, composition of parallel processes.

### 2017-07-05 paper/content/conclusion.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

Arrows turn out to be a useful tool for composing in parallel programs. We do not have to introduce new monadic types that wrap the computation. Instead use arrows just like regular sequential pure functions. 
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows feature an implementation of the |ArrowParallel| interface for Multicore Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with no effort.
%
%
Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.\olcomment{PROOFS. Many proofs in benchmarks!}

\mbcomment{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}

\olcomment{Remove websites from citations, put them into footnotes! --
\done}

\olcomment{Parrows + accelerate = love? Metion port to Frege. Mention the Par monad troubles.}

@@

\olcomment{replace API with DSL globally? \done}

>>>>

Arrows turn out to be a useful tool for composing in parallel programs. We do not have to introduce new monadic types that wrap the computation. Instead use arrows just like regular sequential pure functions. 
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows feature an implementation of the |ArrowParallel| interface for Multicore Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with little to no effort.
%
%
Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.

In order to fix the issues of the Par Monad backend with the |torus| and |ring| skeleton as explained in sections \ref{sec:ring} and \ref{sec:torus} - or any looping skeleton for that matter - additional work that fixes the strictness problems of the |ArrowParallel| instance is required.
\olcomment{Parrows + accelerate = love? Metion port to Frege.}

@@

### 2017-07-05 paper/content/mapSkeletons.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

\paragraph{Statically load-balancing parallel |map|.}
Our |parMap| spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper). This can be quite wasteful and a |farm| (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers seems useful. \olcomment{it is a \emph{static} load balancing}
The definitions of the helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.

>>>>

\paragraph{Statically load-balancing parallel |map|.}
Our |parMap| spawns every single computation in a new thread (at least for the instances of |ArrowParallel| we gave in this paper). This can be quite wasteful and a statically load-balancing |farm| (Figs.~\ref{fig:farmImg},~\ref{fig:farm}) that equally distributes the workload over |numCores| workers seems useful.
The definitions of the helper functions |unshuffle|, |takeEach|, |shuffle| (Fig.~\ref{fig:edenshuffleetc}) originate from an Eden skeleton\footnote{Available on Hackage under \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/src/Control-Parallel-Eden-Map.html}.}.

### 2017-07-05 paper/content/parallelHaskells.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

\subsubsection{ParMonad}
The |Par| monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a monad designed for composition of parallel programs. Let:

@@

The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| monad with |get|. This results in |Par [b]|. We execute this process with |runPar| to finally get the final |[b]|.
A graphical representation can be found Fig.~\ref{fig:parEvalNParMonadImg}.

\mbcomment{explain problems with laziness here. Problems with torus}

@@

While Eden also comes with a monad |PA| for parallel evaluation, it also ships with a completely functional interface that includes
%\\
a |spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]|
%\\
%a |spawnF| 
function that
%This 
allows us to define |parEvalN| directly:

\begin{code}
parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
parEvalN = spawnF 
\end{code}
% \begin{figure}[h]
% 	\includegraphics[scale=0.5]{images/parEvalNEden}
% 	\caption{Dataflow of the Eden |parEvalN| version.}
% 	\label{fig:parEvalNEden}
% \end{figure}
% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.
\olcomment{removed the image as it shows nothing new}

>>>>

\subsubsection{Par Monad}
The |Par| Monad\footnote{It can be found in the \texttt{monad-par} package on hackage under \url{https://hackage.haskell.org/package/monad-par-0.3.4.8/}.} introduced by \citet{par-monad}, is a Monad designed for composition of parallel programs. Let:

@@

The |Par| Monad version of our parallel evaluation function |parEvalN| is defined by zipping the list of |[a -> b]| with the list of inputs |[a]| with the application operator |$| just like with Multicore Haskell. % $
Then, we map over this not yet evaluated lazy list of results |[b]| with |spawnP :: NFData a => a -> Par (IVar a)| to transform them to a list of not yet evaluated forked away computations |[Par (IVar b)]|, which we convert to |Par [IVar b]| with |sequenceA|. We wait for the computations to finish by mapping over the |IVar b| values inside the |Par| Monad with |get|. This results in |Par [b]|. We execute this process with |runPar| to finally get |[b]|.
A graphical representation can be found in Fig.~\ref{fig:parEvalNParMonadImg}.

@@

While Eden also comes with a Monad |PA| for parallel evaluation, it also ships with a completely functional interface that includes
%\\
a |spawnF :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]|
%\\
%a |spawnF| 
function that
%This 
allows us to define |parEvalN| directly:

\begin{code}
parEvalN :: (Trans a, Trans b) => [a -> b] -> [a] -> [b]
parEvalN = spawnF 
\end{code}
% \begin{figure}[h]
% 	\includegraphics[scale=0.5]{images/parEvalNEden}
% 	\caption{Dataflow of the Eden |parEvalN| version.}
% 	\label{fig:parEvalNEden}
% \end{figure}
% A simplistic graphical depiction of this definition can be found in Fig.~\ref{fig:parEvalNEden}.

### 2017-07-05 paper/content/parrows.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}% $ %% formatting
\caption{Fully evaluating |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this. We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
% \begin{figure}[h]
% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}
% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances. To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[t]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}
\caption{Configurable |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{|Par| Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"?}
The |Par| monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| monad's |spawnP :: NFData a => a -> Par (IVar a)| function. The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| monad.

We therefore apply each function to its corresponding input value with and then fork the computation away with |arr spawnP| inside a |zipWithArr| (Fig.~\ref{fig:zipWithArr}) call. This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|. In order to wait for the computation to finish, we map over the |IVar|s inside the |Par| monad with |arr (>>= mapM get)|. The result of this operation is a |Par [b]| from which we can finally remove the monad again by running |arr runPar| to get our output of |[b]|.
\begin{figure}[h]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = 
			(arr $ \as -> (fs, as)) >>>
			zipWithArr (app >>> arr spawnP) >>>
			arr sequenceA >>>
			arr (>>= mapM get) >>>
			arr runPar
\end{code} %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend.}
\label{fig:ArrowParallelParMonad}
\end{figure}

>>>>

instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}% $ %% formatting
\caption{Fully evaluating |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreRdeepseq}
\end{figure}
For most cases a fully evaluating version like in Fig.~\ref{fig:ArrowParallelMulticoreRdeepseq} would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the |Strategy| interface, our DSL should allow for this. We therefore introduce the |Conf a| data-type that simply wraps a |Strategy a|.
% \begin{figure}[h]
% \caption{Definition of Conf a}
% \label{fig:confa}
% \end{figure}
% We can't directly use the |Strategy a| type here as GHC (at least currently) does not allow type synonyms in type class instances. To get our configurable |ArrowParallel| instance, we simply unwrap the strategy and pass it to |parList| like in the fully evaluating version (Fig.~\ref{fig:ArrowParallelMulticoreConfigurable}).
\begin{figure}[t]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b (Conf b) where
    	parEvalN (Conf strat) fs =
        	listApp fs >>>
        	arr (withStrategy (parList strat)) &&& arr id >>>
        	arr (uncurry pseq)
\end{code}
\caption{Configurable |ArrowParallel| instance for the Multicore Haskell backend.}
\label{fig:ArrowParallelMulticoreConfigurable}
\end{figure}
\subsubsection{|Par| Monad}
\olcomment{introduce a newcommand for par-monad, "arrows", "parrows" and replace all mentions to them to ensure uniform typesetting \done, we write Arrows. also "Monad"? \done}
The |Par| Monad implementation (Fig.~\ref{fig:ArrowParallelParMonad}) makes use of Haskells laziness and |Par| Monad's |spawnP :: NFData a => a -> Par (IVar a)| function. The latter forks away the computation of a value and returns an |IVar| containing the result in the |Par| Monad.

We therefore apply each function to its corresponding input value with and then fork the computation away with |arr spawnP| inside a |zipWithArr| (Fig.~\ref{fig:zipWithArr}) call. This yields a list |[Par (IVar b)]|, which we then convert into |Par [IVar b]| with |arr sequenceA|. In order to wait for the computation to finish, we map over the |IVar|s inside the |Par| Monad with |arr (>>= mapM get)|. The result of this operation is a |Par [b]| from which we can finally remove the Monad again by running |arr runPar| to get our output of |[b]|.
\begin{figure}[h]
\begin{code}
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = 
			(arr $ \as -> (fs, as)) >>>
			zipWithArr (app >>> arr spawnP) >>>
			arr sequenceA >>>
			arr (>>= mapM get) >>>
			arr runPar
\end{code} %$ %% formatting
\caption{ArrowParallel instance for the Par Monad backend.}
\label{fig:ArrowParallelParMonad}
\end{figure}

### 2017-07-05 paper/content/relwork.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

\olcomment{arrows or Arrows?}

\paragraph{Parallel Haskells.}
Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work. We use these languages as backends: our DSL can switch from one to other at user's command.

HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of |Par| monad to heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of |Par| monad.
%
Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015}, and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}. Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden. This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.

>>>>

\paragraph{Parallel Haskells.}
Of course, the three parallel Haskell flavours we have presented above: the GpH \cite{Trinder1996,Trinder1998a} parallel Haskell dialect and its multicore version \cite{Marlow2009}, the |Par| Monad \cite{par-monad,Foltzer:2012:MPC:2398856.2364562}, and Eden \cite{eden,Loogen2012} are related to this work. We use these languages as backends: our DSL can switch from one to other at user's command.

HdpH \cite{Maier:2014:HDS:2775050.2633363,stewart_maier_trinder_2016} is an extension of |Par| Monad to heterogeneous clusters. LVish \cite{Kuper:2014:TPE:2666356.2594312} is a communication-centred extension of |Par| Monad.
%
Further parallel Haskell approaches include pH \cite{ph-book}, research work done on distributed variants of GpH \cite{Trinder1996,Aljabri:2013:DIG:2620678.2620682,Aljabri2015}, and low-level Eden implementation \cite{JostThesis,berthold_loidl_hammond_2016}. Skeleton composition \cite{dieterle_horstmeyer_loogen_berthold_2016}, communication \cite{Dieterle2010}, and generation of process networks \cite{Horstmeyer2013} are recent in-focus research topics in Eden. This also includes the definitions of new skeletons \cite{doi:10.1142/S0129626403001380,Eden:PARCO05,Berthold2009-mr,Berthold2009-fft,dieterle2010skeleton,delaEncina2011,Dieterle2013,janjic2013space}.

### 2017-07-05 paper/content/topologySkeletons.tex
mentioning Par Monad problems now. "Monad" instead of "monad"

(parcomp) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
	arr a b -> arr b c -> arr a c
(parcomp) = pipe2 ()
\end{code}
\caption{Definition of |pipe2| and a parallel |>>>|.}
\label{fig:pipe2}
\end{figure}
\mbcomment{parallel >>>!!!!!!}

@@

Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into a our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|. Finally we unzip this list into |([o], [fut r])|. Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
%Further details on this can be found in \cite{eden_cefp}.
\olcomment{let's do it?}

@@

\subsection{Torus skeleton}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the |torus| skeleton.}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a ring from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.

@@

>>>>

(parcomp) :: (ArrowLoop arr, ArrowChoice arr, Future fut (([a], [b]), [c]),
	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) ()) =>
	arr a b -> arr b c -> arr a c
(parcomp) = pipe2 ()
\end{code}
\caption{Definition of |pipe2| and a parallel |>>>|.}
\label{fig:pipe2}
\end{figure}

@@

Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)| so we can feed that into a our input arrow |arr (i, r) (o, r)|, which we transform into |arr (i, fut r) (o, fut r)| before lifting it to |arr [(i, fut r)] [(o, fut r)]| to get a list |[(o, fut r)]|. Finally we unzip this list into |([o], [fut r])|. Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}).
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.

Note that the Par Monad version of the |ArrowParallel| version  presented in this paper (Fig.~\ref{fig:ArrowParallelParMonad}) does not work with the looping behaviour in the |ring| skeletons. Whenever |runPar| is invoked, strictness is enforced, which means that the |lazy| operation in the definition of |ring| will not work.
%Further details on this can be found in \cite{eden_cefp}.
\olcomment{let's do it?}

@@

\subsection{Torus skeleton}\label{sec:torus}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the |torus| skeleton.}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a ring from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.

@@

Again, the implementation of the Par Monad version prevents this definition of the |torus| skeleton to work due to the reasons explained in the section about the |ring| skeleton in section \ref{sec:ring}.
